From Questions to Queries – Enhancing the Curt System by a Theory and Implementation of Embedded WH-Questions

### 2008-08-29 thesis/paper.tex
Some corrections to the the first two sections.

Development of exhaustive systems in this area appears to be a daunting task of
quite some scope. Therefore most approaches have been focused on providing
coverage of a small fragment of natural language often with respect to a
particular domain. Many such systems exist and they prove to be increasingly
successful. Some of them will be discussed here and presenting an extension to
one system in particular is the main matter of this Bachelor's thesis.

@@

The present work will aim to provide the \curt{} system of
\cite{blackburnbos:cl1} with the mechanics of dealing with more than one
partition of its logical domain, viz. the ability to handle more than one
\term{possible} or \term{accessible world}. As a ground for development of this
work, the treatment of \term{embedded questions} was chosen as one that
naturally demands a possibility to account for more than one division in a
system's \emph{understanding} or representation of a universe of objects and
relations.

@@

Computational Semantics is a discipline that has  lately gained a wider focus in
literature due to recent progress in different areas.  Current development in
computer hardware has made it possible to focus on theoretical soundness and
important abstract designs and concepts of logic and the theory of meaning
rather than being able to simply implement a system that is able to do basic
computations in a reasonable amount of time. On the other hand, research in both
computer science and linguistics has fueled new ideas and intriguing hypotheses
emerge from lively discussion in their respective fields on a very frequent
basis.

This section delineates the ground the software presented in this thesis is
based on. \ref{sec:curt} introduces the base system of the implementation,
\curt{}
and \ref{sec:comparison} relates it to other similar systems in existence.
\ref{sec:extension} then goes on to explain where the \curt{} system lacks in
features and thus exhibits potential for extension.

@@

The final chapter of \cite{blackburnbos:cl1} introduces the implementation
incrementally, presenting new versions of \curt{} that add to the software's
feature set one at a time. Support for inferencing via external theorem provers
and model builders is added right after the first, rather limited implementations.
Finally an ontology and a straight-forward way of parsing and answering direct
questions for noun phrases conclude its development. The theory and
implementation of this paper is meant as a replacement and enhancement for the
latter feature. This will also happen in an incremental manner and new features
are introduced one at a time during section \ref{sec:implementation}.

@@

Interestingly, earlier attempts at providing a software capable of answering
questions in natural language where also mostly coupled with the usage of
semantic analysis instead of statistical methods.
\cite{chat} are among  the first to attempt creating a question answering system
with reasonable semantic coverage. Their effort has been focused on the geographical
domain, and their implementation, dubbed \pn{CHAT-80}, is written entirely in
\prol. The lack of hardware resources of that time forced the system to be
written with efficiency in mind. Moreover, it had to maintain a statical
database of facts which was hand-crafted. As such, it falls more into the
category of question answering systems than dialogue systems.

Today it seems that question answering systems make inroads to the search market,
and some argue that they are bound to eventually supplant current document/term-based
search engines. Even though \abbr{QA} systems also operate on the domain of
answering questions, their approach to resolving them differs in most aspects
from that of dialogue systems. The latter have a much higher obligation to
correctness, since their results are typically not subject to peer
review and server an entirely different purpose.

@@

\curt{} is a dialogue system and thus based on an entirely different set of
premises than \abbr{QA} systems. Moreover, it differs from \pn{Godot} in some
major points, the most distinctive of which is that \curt{} only exists inside a
regular computer and has no access to sensory input. Also, \curt's
representation of knowledge differs from that of \pn{Godot} or the \pn{Pronto QA
System} in that it does not use discourse representation theory (yet). \acurt{}
addresses some of its shortcomings and lays ground to the implementation of a
broader feature set for it.

@@

In the course of this thesis, a new version of \curt{} will be developed, called
\acurt. This version is not only aware of its own state, but can also gather
information about the individuals it knows about. Furthermore, \acurt{} is able
to account for several possible readings of a certain sentence at once and in
parallel, which is a side-effect of the introduction of the notion of
\term{possible worlds} into the system.

@@

For this to work, the internal storage of \curt{} will have to be restructured
significantly in order to deal with more than one possible world or situation.
Its grammar will have to be enhanced so that it can handle \wh-embedding verbs
like `know' and `believe' and its inferencing mechanisms are going to be
employed in a different manner. %todo sucks.

@@

Question semantics has traditionally been a very sophisticated field of
linguistics and philosophy, with many different approaches arguing for and
against each other. This section will not try to give an exhausting overview of
the field of question semantics since this would be a task of considerable size
and by far beyond the scope of this thesis\footnote{For a good summary of some
of the literature, particularly the issue of whether questions are a concern of
pragmatics or semantics, consider \cite[section 2 and 3]{gs:q}.}. Instead, it
was chosen to roughly follow the work by Groenendijk and Stokhof, lending from
other places where deemed appropriate or practical. As such, the theoretical
framework of \acurt{} is not going to be an exact realization of any particular
approach to question semantics, but rather an accumulation of interesting
concepts from different sources. The main sources, however, are \cite{gs:q} and
\cite{gs:sqpa}.

@@

Note, however, that all choices presented in this section have been made
with the implementation of section \ref{sec:implementation} in mind.
Sometimes, theoretical soundness might suffer from these choices as the main
goal of the present work is to provide a proof-of-concept of a software that is
able to reason about possible worlds, employing first-order logic inferencing
tools. % really? fix your title then, would you? … doh. fixme

@@

Following \cite{gs:q}, we will not use the English word
``question'' in its traditional sense, but distinguish between \term{questions},
\term{interrogative sentences} and \term{interrogative acts} in the following
way:

\begin{termin}
  An \term{interrogative} is a certain type of sentence in natural language
  distinguished by certain characteristics
  \footnote{To quote \cite{gs:q}: \quotesf{[An
  interrogative sentence is] characterized by word order, intonation, question
  mark, the occurrence of interrogative pronouns}.}
  whereas an \term{interrogative act} refers to the utterance of an
  \term{interrogative}. Finally, a
  \term{question} will denote the semantic entity uttered by the
  \term{interrogative act}, as it will be introduced in section
  \ref{sec:protoq}
\end{termin}

@@

The following abbreviations will be used: 
\abbr{Ty2} stands for two-sorted type logic\footnote{See \cite{gallin:ty2} for a
discussion and \cite{gs:sqpa} for its usage in describing the semantics of
questions}. \abbr{IL} stands for intensional logic\footnote{As used by
\cite{ptq}. See also \cite{gamut:2} for an excellent introduction to \abbr{IL}.},
\abbr{FOL}  is used as an abbreviation for first order logic.

@@

Most of the traditional literature in natural language semantics follows
\cite{ptq} and thus also makes use of its underlying formalism,
\abbr{IL}. \cite{gs:sqpa} however use \abbr{Ty2} as it was
proposed in \cite{gallin:ty2}. \cite{z:ilty2} discusses the differences and
similarities in expressive power between \abbr{Ty2} and \abbr{IL} and comes to
the conclusion that the set of expressions contained in the former but not in
the latter might not be relevant for the treatment of natural language
semantics. Although he proposes a translation scheme between both languages,
translation from \abbr{Ty2} to \abbr{IL} seems to sometimes produce very complex
\abbr{IL} terms.

As was already mentioned in \ref{sec:curt}, \curt{} is based on 
the use of external tools only capable of treating first order logic – an
extensional framework. \prol{} on the other hand, is a language that can very well
make use of higher order formalisms for calculation\footnote{The two most
popular predicates in \prol{} used for implementing higher-order logic are
\code{call/n} and \code{apply/3}. This paper is based on \pn{SWI-Prolog} which
provides an implementation of the former, but not of the latter. In addition
\pn{SWI-Prolog} also defines several higher order aggregate functions, including
\code{maplist/4}, which will be used later on. See \cite{naish:prolhio} for a
discussion of higher order functional programming in \prol.}.  While there exist
inferencing tools for modal logic\footnote{ The interested reader is referred to
\pn{Molle} (\url{http://molle.sf.net}), a modal theorem prover that comes with a
graphical user interface which allows interactive investigation of the
constructed indices.}, a restructuring of the \curt{} system to take advantage
of them would be beyond the scope of this thesis.  It was therefore decided to
avoid modal logic and implement a pragmatic approach to representing the indices
of higher-order logic.

Some research on the logic of questions (eroteric logic) suggests that the
treatment of interrogatives demands its own formalism.  \acurt{} employs a
formalism that departs from both \abbr{IL} and \abbr{Ty2} in some ways. Most
notably, the notion of a \term{model} is changed as follows, in order to
accommodate the question operator introduced further on, which demands a set of
possible worlds to be available.

\begin{definition}\label{Models}
  \stress{Models\footnote{See also the Appendix, section \ref{sec:examplemodel}
  for an example of such a model.}}
  \begin{enumerate}
    \item A model is a set of pairs of \term{indices} $i$ and \term{possible
    worlds} $w$: $\pair{i,w} \in M$. 
    \item A world $w$ at index $i$, $\pair{i,w} \in M$ is a triple of $D$, the set
    of individuals, $F$, the set of functions, and $\varphi$ a first-order
    logical formula that holds in this particular model: $w = \pair{D,F,\varphi}$
    \item A \term{function} $f \in F$ is one of the following: 
    \begin{itemize}
      \item $f(a,\sigma,L)$, where $\sigma$ is the function symbol, $a$ denotes
      the function symbol's arity and $L$, the set of arguments, for which the
      following holds: $|L| = a$,
      $L = \set{\pair{x_1,\ldots{},x_a}|x_1,\ldots{},x_a \in D}$
      \item The special case $f(0,\sigma,c)$, where $c \in D$, which denotes the
      assignment of a proper name $\sigma$ to $c$.
    \end{itemize}
    \item $\pair{i,w}, \pair{j,w'} \in M, w = w'$ iff $w = \pair{D,F,\varphi}, w' =
    \pair{D',F',\varphi'}$ and $\varphi = \varphi'$.
    \item If $W \in M$, then we say $\forall W$ quantifies over all members of $M$.
    \item If $W \in M$, then we say $\exists W$ quantifies over the existence of
    one member of $M$, namely $|M| \geq 1$ and $W \in M$.
  \end{enumerate}
\end{definition}

One of the more important differences to the `traditional' notion of a model, is
that it was decided to assign every world a unique domain of discourse referents
or \term{individuals}. Note that there is no requirement $\pair{i,\pair{D,F}},
\pair{j,\pair{D',F'}} \in M \wedge D\neq D'$, and most of the time, a model's
possible worlds will indeed share many of their individuals or have entirely
equal domains, but diversion in the world's domains will sometimes be useful and
unavoidable.

@@

\ex. \label{ex:ginzburg}
\a. Jill believed a certain hypothesis. Hence, Jill believed that hypothesis
to be true.
\b. Jill discovered a certain hypothesis. It \emph{does not follow} that Jill
discovered that hypothesis to be true. \cite{ginzburg} defends an analysis that
introduces another distinction in the nature of indicatives: the distinction
between \term{facts} and \term{propositions} -- where in \ref{ex:ginzburg} the
embedded clause is treated as a proposition and not as a fact.

@@

  Every other operator or function can take a set of propositions of the type of
  $\rho'$ to produce a set of propositions containing \emph{all possible
  permutations} of $\rho'$. For every possible operator $\sem{$\cdot$}$ with arity
  $n$ and arguments $\sem{$\cdot$}(\set{\varphi,\psi})$

@@

  \begin{eqnarray*}
    \sem{$\cdot$}(\rho') \mapsto \set{\varphi,\psi}\\
    \sem{$\cdot$}(\sigma_1,\dots,\sigma_j,\rho )

    \sigma \wedge \set{\varphi, \psi} \mapsto \set{\sigma\wedge\varphi,\sigma\wedge\psi}\\
    \sigma \to \set{\varphi, \psi} \mapsto \set{\sigma\to\varphi,\sigma\to\psi}\\
  \end{eqnarray*}
\end{definition}

>>>>

Development of exhaustive systems in this area appears to be a daunting task of
quite some scope. Therefore most approaches have been focused on providing
coverage of a small fragment of natural language, often with respect to a
particular domain. Many such systems exist and they prove to be increasingly
successful. Some of them will be discussed here, and presenting an extension to
one system in particular is the main matter of this Bachelor's thesis.

@@

The present work will aim to provide the \curt{} system of
\cite{blackburnbos:cl1} with the mechanics of dealing with more than one
partition of its logical domain, viz. the ability to handle more than one
\term{possible} or \term{accessible world}. As a ground for the development of this
work, the treatment of \term{embedded questions} was chosen as one that
naturally demands a possibility to account for more than one division in a
system's \emph{understanding} or representation of a universe of objects and
relations.

@@

Computational Semantics is a discipline that has  lately gained a wider focus in
literature due to recent progress in different areas.  Current development in
computer hardware has made it possible to focus on theoretical soundness,
important abstract designs, and concepts of logic and the theory of meaning,
rather than being able to simply implement a system that is able to do basic
computations in a reasonable amount of time. On the other hand, research in both
computer science and linguistics has fueled new ideas and intriguing hypotheses
emerge from lively discussions in their respective fields on a very frequent
basis.

This section delineates the ground the software presented in this thesis is
based on. \ref{sec:curt} introduces the base system of the implementation,
\curt{},
and \ref{sec:comparison} relates it to other similar systems in existence.
\ref{sec:extension} then goes on to explain where the \curt{} system lacks in
features and thus exhibits potential for extension.

@@

The final chapter of \cite{blackburnbos:cl1} introduces the implementation
incrementally, presenting new versions of \curt{} that add to the software's
feature set one at a time. Support for inferencing via external theorem provers
and model builders is added right after the first, rather limited implementations.
Finally an ontology and a straight-forward way of parsing and answering direct
questions for noun phrases conclude its development. The theory and
implementation of this paper are meant as a replacement and enhancement for the
latter feature. Their introduction will also happen in an incremental manner and new features
are introduced one at a time during section \ref{sec:implementation}.

@@

Interestingly, earlier attempts at providing a software capable of answering
questions in natural language were mostly coupled with the usage of
semantic analysis instead of statistical methods.
\cite{chat} are among  the first to attempt creating a question answering system
with reasonable coverage. Their effort has been focused on the geographical
domain, and their implementation, dubbed \pn{CHAT-80}, is written entirely in
\prol. The lack of hardware resources of that time forced the system to be
written with efficiency in mind. Moreover, it had to maintain a statical
database of facts which was hand-crafted. As such, it falls more into the
category of question answering systems than dialogue systems.

Today it seems that statistical question answering systems make inroads to the search market,
and some argue that they are bound to eventually supplant current document/term-based
search engines. Even though \abbr{QA} systems also operate on the domain of
answering questions, their approach to resolving them differs in most aspects
from that of dialogue systems. The latter have a much higher obligation to
correctness, since their results are typically not subject to peer
review and serve an entirely different purpose.

@@

\curt{} is a prototype dialogue system and as such it is based on an entirely different set of
premises than \abbr{QA} systems. Moreover, it differs from \pn{Godot} in some
major points, the most distinctive of which is that \curt{} only exists inside a
regular computer and has no access to sensory input. Also, \curt's
representation of knowledge differs from that of \pn{Godot} or the \pn{Pronto QA
System} in that it does not use discourse representation theory (yet). \acurt{}
addresses some of its shortcomings and lays ground to the implementation of a
broader feature set for it.

@@

In the course of this thesis, a new version of \curt{} is introduced, called
\acurt. This version is not only aware of its own state of knowledge, as was the
original \curt{} implementation, but can also gather information about the
knowledge and beliefs of individuals it knows about. Furthermore, \acurt{} is
able to account for several possible readings of a certain sentence at once and
in parallel, which is a side-effect of the introduction of the notion of
\term{possible worlds} into the system.

@@

For this to work, the internal storage of \curt{} will have to be restructured
significantly in order to deal with more than one possible world or situation.
Its grammar will have to be enhanced so that it can handle \wh-embedding verbs
like `know' and `believe' and its inferencing mechanisms are going to be
employed in a different manner, to encompass a wider range of functionality.

@@

The semantics of questions has traditionally been a very sophisticated field of
linguistics,  philosophy and logics, with many different approaches arguing for
and against each other. This section will not try to give an exhausting overview
of the field of question semantics since this would be a task of considerable
size and by far beyond the scope of this thesis\footnote{For a good summary of
some of the literature, particularly the issue of whether questions are a
concern of pragmatics or semantics, consider \cite[section 2 and 3]{gs:q}.}.
Instead, it was chosen to roughly follow the work by Groenendijk and Stokhof,
who present one of the best-established and recognized theories in the field,
and also lending from other places where deemed appropriate or practical. As
such, the theoretical framework of \acurt{} is not going to be an exact
realization of any particular approach to question semantics, but rather an
accumulation of interesting concepts from different sources. The main sources,
however, are \cite{gs:q} and \cite{gs:sqpa}.

@@

Note, however, that all choices presented in this section have been made
with the implementation of section \ref{sec:implementation} in mind.
Sometimes, theoretical soundness might suffer from these choices as the main
goal of the present work is to provide a proof-of-concept of a software that is
able to reason about possible worlds, employing first-order logic inferencing
tools in order to realize an approach to the treatment of embedded questions and
questions in general.

@@

Following \cite{gs:q}, the English word ``question'' is not used in its
traditional sense, but a distinction is made between \term{questions},
\term{interrogative sentences} and \term{interrogative acts} in the following
way:

\begin{termin}
  An \term{interrogative sentence}, or simply \term{interrogative}, is a
  type of sentence in natural language
  distinguished by certain characteristics
  \footnote{To quote \cite{gs:q}: \quotesf{[An
  interrogative sentence is] characterized by word order, intonation, question
  mark, the occurrence of interrogative pronouns}.},
  whereas an \term{interrogative act} refers to the utterance of an
  \term{interrogative}. Finally, a
  \term{question} will denote the semantic entity uttered by the
  \term{interrogative act}, as it will be introduced in section
  \ref{sec:protoq}
\end{termin}

@@

The following abbreviations will be used: 
\abbr{Ty2} stands for two-sorted type logic\footnote{See \cite{gallin:ty2} for a
discussion and \cite{gs:sqpa} for its usage in describing the semantics of
questions}, \abbr{IL} stands for intensional logic\footnote{As used by
\cite{ptq}. See also \cite{gamut:2} for an excellent introduction to \abbr{IL}.},
\abbr{FOL}  is used as an abbreviation for first order logic. As an alternative
term for \term{possible world}, sometimes the expression \term{index} is used.

@@

Most of the traditional literature in natural language semantics follows
\cite{ptq} and thus also makes use of its underlying formalism,
\abbr{IL}. \cite{gs:sqpa}, however, use \abbr{Ty2} as it was
proposed in \cite{gallin:ty2}. \cite{z:ilty2} discusses the differences and
similarities in expressive power between \abbr{Ty2} and \abbr{IL} and comes to
the conclusion that the set of expressions contained in the former but not in
the latter might not be relevant for the treatment of natural language
semantics. Although he proposes a translation scheme between both languages,
translation from \abbr{Ty2} to \abbr{IL} seems to sometimes produce very complex
\abbr{IL} terms.

As was already mentioned in \ref{sec:curt}, \curt{} is based on 
the use of external tools. These tools are only capable of treating first order logic – an
extensional framework. \prol{} on the other hand, is a language that can very well
make use of higher order formalisms for calculation\footnote{The two most
popular predicates in \prol{} used for implementing higher-order logic are
\code{call/n} and \code{apply/3}. This paper is based on \pn{SWI-Prolog} which
provides an implementation of the former, but not of the latter. In addition
\pn{SWI-Prolog} also defines several higher order aggregate functions, including
\code{maplist/4}, which will be used later on. See \cite{naish:prolhio} for a
discussion of higher order functional programming in \prol.}.  While there exist
inferencing tools for modal logic\footnote{ The interested reader is referred to
\pn{Molle} (\url{http://molle.sf.net}), a modal theorem prover that comes with a
graphical user interface which allows interactive investigation of the
constructed indices.}, a restructuring of the \curt{} system to take advantage
of them would be beyond the scope of this thesis.  It was therefore decided to
avoid modal logic and implement a pragmatic approach to representing indices in
order to increase the expressive power of the system.

Some research on the logic of questions (sometimes also referred to as eroteric
logic) suggests that the treatment of interrogatives demands its own formalism.
\acurt{} employs a formalism that departs from both \abbr{IL} and \abbr{Ty2} in
some ways. Most notably, the notion of a \term{model} is changed as follows, in
order to accommodate the question operator introduced further on, which demands
a set of possible worlds to be available.

\begin{definition}\label{Models}
  \stress{Models\footnote{See also the Appendix, section \ref{sec:examplemodel}
  for an example of such a model.}}
  \begin{enumerate}
    \item A model $M$ is a set of pairs of \term{indices} $i$ and \term{possible
    worlds} $w$: $\pair{i,w} \in M$. 
    \item A world $w$ at index $i$, $\pair{i,w} \in M$, is a triple of $D$, the set
    of individuals, $F$, the set of functions, and $\varphi$ a first-order
    logical formula that holds in this particular model: $w = \pair{D,F,\varphi}$
    \item A \term{function} $f \in F$ is one of the following: 
    \begin{itemize}
      \item $f(a,\sigma,L)$, where $\sigma$ is the function symbol, $a$ denotes
      the function symbol's arity and $L$, the set of arguments, for which the
      following holds: $|L| = a$,
      $L = \set{\pair{x_1,\ldots{},x_a}|x_1,\ldots{},x_a \in D}$
      \item The special case $f(0,\sigma,c)$, where $c \in D$, which denotes the
      assignment of a proper name $\sigma$ to $c$.
    \end{itemize}
    \item $\pair{i,w}, \pair{j,w'} \in M, w = w'$ iff $w = \pair{D,F,\varphi}, w' =
    \pair{D',F',\varphi'}$ and $\varphi = \varphi'$.
    \item If $W \in M$, then we say $\forall W$ quantifies over all members of $M$.
    \item If $W \in M$, then we say $\exists W$ quantifies over the existence of
    one member of $M$, so $|M| \geq 1$ and $W \in M$.
  \end{enumerate}
\end{definition}

One of the more important differences to the `traditional', \abbr{FOL}, notion of a model, is
that it was decided to assign every world a unique domain of discourse referents
or \term{individuals}. Note that there is no requirement $\pair{i,\pair{D,F}},
\pair{j,\pair{D',F'}} \in M \wedge D\neq D'$, and most of the time, a model's
possible worlds will indeed share many of their individuals or have entirely
equal domains, but diversion in the world's domains will sometimes be useful and
unavoidable.

@@

\ex. \label{ex:ginzburg}
\a. Jill believed a certain hypothesis. Hence, Jill believed that hypothesis
to be true.
\b. Jill discovered a certain hypothesis. It \emph{does not follow} that Jill
discovered that hypothesis to be true.

\cite{ginzburg} defends an analysis that
introduces another distinction in the nature of indicatives: the distinction
between \term{facts} and \term{propositions} -- where in \ref{ex:ginzburg} the
embedded clause is treated as a proposition and not as a fact.

@@

  %Every other operator or function can take a set of propositions of the type of
  %$\rho'$ to produce a set of propositions containing \emph{all possible
  %permutations} of $\rho'$. For every possible operator $\sem{$\cdot$}$ with arity
  %$n$ and arguments $\sem{$\cdot$}(\set{\varphi,\psi})$

@@

  \begin{eqnarray*}
    %\sem{$\cdot$}(\rho') \mapsto \set{\varphi,\psi}\\
    %\sem{$\cdot$}(\sigma_1,\dots,\sigma_j,\rho )
    \varphi
    %\sigma \wedge \set{\varphi, \psi} \mapsto \set{\sigma\wedge\varphi,\sigma\wedge\psi}\\
    %\sigma \to \set{\varphi, \psi} \mapsto \set{\sigma\to\varphi,\sigma\to\psi}\\
  \end{eqnarray*}
\end{definition}
